---
title: "Getting Started with CaImagingAnalysisFr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with CaImagingAnalysisFr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = FALSE
)
```

This vignette introduces the basic workflow for analyzing calcium imaging data with CaImagingAnalysisFr.

## Installation

```{r install}
# Install from GitHub
# install.packages("remotes")
remotes::install_github("kevinj24fr/CaImagingAnalysisFr")
```

## Loading the Package

```{r load}
library(CaImagingAnalysisFr)
```

## Basic Workflow

A typical calcium imaging analysis pipeline consists of:

1. Loading data
2. Motion correction
3. Cell segmentation / ROI extraction
4. Trace extraction and preprocessing
5. Spike inference
6. Downstream analysis

### Step 1: Generate or Load Data

For this tutorial, we'll use synthetic data:

```{r data}
# Generate synthetic calcium imaging data
set.seed(42)
data <- generate_synthetic_data(
 n_cells = 50,
 n_time = 3000,
 spike_prob = 0.02,
 frame_rate = 10
)

# View structure
dim(data)
```

For real data, you can load TIFF stacks:

```{r load-tiff}
# Load a TIFF stack
movie <- read_tiff_stack("path/to/recording.tif")

# Or load from various formats (HDF5, NWB, MATLAB)
movie <- read_calcium_imaging("path/to/data.h5", format = "hdf5")
```

### Step 2: Motion Correction

Correct for sample movement during recording:

```{r motion}
# Apply motion correction
corrected_movie <- motion_correct(
 movie,
 method = "template",
 max_shift = 20
)

# Assess correction quality
quality <- assess_motion_correction(movie, corrected_movie)
print(quality$improvement)
```

### Step 3: Cell Segmentation

Identify individual cells in the field of view:

```{r segment}
# Segment cells using watershed method
masks <- segment_cells(
 corrected_movie,
 method = "watershed",
 min_area = 50,
 max_area = 500
)

# Get ROI properties
props <- region_properties(masks)
print(paste("Found", nrow(props), "cells"))
```

### Step 4: Extract and Preprocess Traces

Extract fluorescence traces from ROIs:

```{r traces}
# Extract raw traces
raw_traces <- extract_traces_from_rois(corrected_movie, masks)

# Apply background correction and compute dF/F
corrected <- calcium_correction(raw_traces, method = "modern")

# Optional: neuropil correction
corrected <- neuropil_correct(
 raw_traces,
 neuropil_traces,
 coefficient = 0.7
)
```

### Step 5: Spike Inference

Detect neural activity events:

```{r spikes}
# OASIS deconvolution (recommended)
spikes <- infer_spikes(corrected, method = "oasis")

# Alternative: threshold-based detection
spikes_threshold <- infer_spikes(corrected, method = "threshold", threshold = 2.5)

# Alternative: Bayesian inference
spikes_bayes <- infer_spikes(corrected, method = "bayesian")
```

### Step 6: Downstream Analysis

#### Functional Connectivity

```{r connectivity}
# Compute pairwise correlations
connectivity <- functional_connectivity(corrected, method = "correlation")

# Graph-theoretic analysis
metrics <- graph_metrics(connectivity)
print(metrics$clustering_coefficient)

# Community detection
communities <- community_detection(connectivity, method = "louvain")
```

#### Stimulus-Aligned Analysis

```{r stimulus}
# Define stimulus times (in frames)
stimulus_times <- c(100, 500, 900, 1300)

# Align traces to stimuli
aligned <- align_to_events(
 traces = corrected,
 events = stimulus_times,
 pre_frames = 30,
 post_frames = 60
)

# Compute response metrics
metrics <- compute_response_metrics(
 aligned,
 baseline_window = c(1, 30),
 response_window = c(31, 60)
)

# Statistical testing
stats <- test_event_responses(aligned, test = "wilcox")
```

## Visualization

```{r viz}
# Plot traces for multiple cells
plot_multiple_cells(corrected, cells = 1:5, frame_rate = 10)

# Heatmap of all cell activity
calcium_activity_heatmap(corrected)

# Network visualization
network_visualization(connectivity, communities = communities$membership)
```

## Quality Control

Generate a comprehensive QC report:

```{r qc}
qc_report <- generate_qc_report(
 traces = corrected,
 spikes = spikes,
 rois = masks
)

print(qc_report)
```

## Next Steps

- See `vignette("pharmacology-analysis")` for drug treatment experiments
- See `vignette("performance-features")` for S7 classes, data.table, and parallel processing

## Session Info

```{r session, eval = TRUE}
sessionInfo()
```
