---
title: "Performance Features and R-Native Tools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance Features and R-Native Tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = FALSE
)
```

CaImagingAnalysisFr includes several R-native performance features for handling large datasets efficiently. This vignette covers S7 classes, data.table operations, parallel processing, and out-of-core computing with Arrow/DuckDB.

## Installation of Optional Dependencies

These features require optional packages:

```{r install-deps}
# S7 type system
install.packages("S7")

# High-performance data operations
install.packages("data.table")

# Parallel processing
install.packages(c("future", "furrr"))

# Out-of-core processing
install.packages(c("arrow", "duckdb"))

# C++ acceleration (requires compilation)
install.packages(c("Rcpp", "RcppArmadillo"))
```

## S7 Type System

S7 provides type-safe classes for calcium imaging data with validation and method dispatch.

### Creating S7 Objects

```{r s7-create}
library(CaImagingAnalysisFr)

# Create a CalciumTraces object
traces <- CalciumTraces(
  data = trace_matrix,      # cells x time matrix
  frame_rate = 10,
  cell_ids = paste0("cell_", 1:nrow(trace_matrix))
)

# Create a CalciumMovie object
movie <- CalciumMovie(
  data = movie_array,       # height x width x time array
  frame_rate = 30
)

# Create an ROI set
rois <- ROISet(
  masks = mask_list,
  centroids = centroid_matrix,
  areas = area_vector
)
```

### Type Validation

S7 objects validate inputs automatically:

```{r s7-validation}
# This will error if frame_rate is negative
traces <- CalciumTraces(data = mat, frame_rate = -10)
# Error: frame_rate must be positive

# This will error if data is not a matrix
traces <- CalciumTraces(data = "not a matrix", frame_rate = 10)
# Error: data must be a numeric matrix
```

### Converting Existing Data

```{r s7-convert}
# Convert matrix to CalciumTraces
traces <- as_calcium_traces(existing_matrix, frame_rate = 10)

# Convert array to CalciumMovie
movie <- as_calcium_movie(existing_array, frame_rate = 30)

# Check if object is a calcium type
is_calcium_object(traces)  # TRUE
```

### Experiment Sessions

Bundle related data together:

```{r s7-session}
session <- ExperimentSession(
  traces = traces,
  spikes = spike_matrix,
  rois = rois,
  metadata = list(
    animal_id = "mouse_001",
    date = "2026-01-15",
    condition = "control"
  )
)

# Access components
session@traces
session@metadata$condition
```

## data.table Operations

data.table provides fast operations on large trace datasets.

### Converting to data.table Format

```{r dt-convert}
# Convert matrix to long-format data.table
dt <- traces_to_dt(trace_matrix, frame_rate = 10)

head(dt)
#    cell_id frame     time    value
# 1:       1     1 0.000000 0.234521
# 2:       1     2 0.100000 0.245123
# ...

# Convert back to matrix
mat <- dt_to_traces(dt)
```

### Fast dF/F Computation

```{r dt-dff}
# Compute dF/F with rolling baseline
dt_dff <- dt_compute_dff(
  dt,
  baseline_frames = 1:100,
  method = "rolling",
  window = 500
)

# Z-score normalization
dt_zscore <- dt_zscore(dt)
```

### Event Detection

```{r dt-events}
# Detect events per cell
events <- dt_detect_events(dt, threshold = 2.5, min_duration = 3)

head(events)
#    cell_id start_frame end_frame peak_frame peak_value
# 1:       1         145       152        148   3.245123
```

### Grouped Operations

```{r dt-grouped}
# Summary statistics per cell
summary <- dt_cell_summary(dt)

head(summary)
#    cell_id     mean       sd      min      max n_events
# 1:       1  0.00234  0.45123 -1.23456  4.56789       12

# Bin traces temporally
dt_binned <- dt_bin_traces(dt, bin_size = 10)

# Apply custom transformation
dt_transformed <- dt_transform(dt, function(x) (x - mean(x)) / sd(x))
```

### Pairwise Correlations

```{r dt-cor}
# Fast pairwise correlation matrix
cor_mat <- dt_pairwise_cor(dt)
```

## Parallel Processing

Use multiple CPU cores for computationally intensive operations.

### Setup

```{r parallel-setup}
# Configure parallel backend
setup_parallel(workers = 8)

# Check current configuration
parallel_info()
# Backend: multisession
# Workers: 8
# Available cores: 16

# When done, shut down workers
# shutdown_parallel()
```

### Parallel Trace Analysis

```{r parallel-traces}
# Apply function to each cell in parallel
results <- parallel_apply_traces(
  traces,
  function(x) {
    # Complex per-cell computation
    list(
      mean = mean(x),
      peaks = find_peaks(x),
      spectrum = compute_power_spectrum(x)
    )
  }
)
```

### Parallel Spike Detection

```{r parallel-spikes}
# Detect spikes across all cells in parallel
spikes <- parallel_spike_detection(
  traces,
  method = "oasis",
  frame_rate = 10
)
```

### Parallel Motion Correction

```{r parallel-motion}
# Process multiple recordings in parallel
corrected_movies <- parallel_motion_correct(
  movie_list,
  method = "template",
  max_shift = 20
)
```

### Batch Processing

```{r parallel-batch}
# Process multiple files
file_list <- list.files("data/", pattern = "*.tif", full.names = TRUE)

results <- parallel_batch_process(
  file_list,
  function(file) {
    movie <- read_tiff_stack(file)
    corrected <- motion_correct(movie)
    traces <- extract_traces(corrected, masks)
    spikes <- infer_spikes(traces)
    list(traces = traces, spikes = spikes)
  }
)
```

## Arrow/DuckDB Backend

For datasets too large to fit in memory, use Arrow and DuckDB for out-of-core processing.

### Saving to Parquet

```{r arrow-save}
# Save traces to Parquet format (compressed, columnar)
save_traces_parquet(traces, "experiment_traces.parquet")

# Load back (memory-mapped, not fully loaded)
traces_arrow <- load_traces_parquet("experiment_traces.parquet")
```

### SQL Queries on Traces

```{r duckdb-query}
# Open DuckDB connection to trace data
con <- duckdb_traces("experiment_traces.parquet")

# Query specific cells or time ranges
active_cells <- query_traces(
  con,
  "SELECT cell_id, AVG(value) as mean_activity
   FROM traces
   WHERE frame BETWEEN 1000 AND 2000
   GROUP BY cell_id
   HAVING AVG(value) > 0.5
   ORDER BY mean_activity DESC"
)

# Don't forget to close
close_duckdb_traces(con)
```

### Streaming Processing

Process data in chunks without loading everything:

```{r arrow-stream}
# Stream process large dataset
results <- stream_process_traces(
  "large_experiment.parquet",
  chunk_size = 10000,
  function(chunk) {
    # Process each chunk
    summarize_chunk(chunk)
  }
)
```

### Partitioned Datasets

For multi-session experiments:

```{r arrow-partitioned}
# Create partitioned dataset (one file per session)
create_partitioned_dataset(
  trace_list,
  "multi_session_data/",
  partition_by = "session_id"
)

# Open partitioned dataset
dataset <- open_partitioned_dataset("multi_session_data/")

# Query across all sessions
all_summaries <- arrow_cell_summary(dataset)
```

## Formula Interface

Use R formulas for common analyses:

```{r formula}
# Analyze traces with formula syntax
results <- analyze_traces(
  dff ~ time | cell_id,
  data = trace_data
)

# Fit response model
model <- fit_response_model(
  amplitude ~ stimulus * condition | cell_id,
  data = response_data
)

# Spike detection with formula
spikes <- detect_spikes_formula(
  spikes ~ trace + threshold(2.5) + method("oasis"),
  data = traces
)
```

### Analysis Pipelines

Define reusable pipelines:

```{r pipeline}
# Define pipeline
pipeline <- define_pipeline(
  result ~ motion_correct(method = "template") +
           extract_traces(masks) +
           calcium_correction(method = "modern") +
           infer_spikes(method = "oasis")
)

# Run on data
results <- run_pipeline(pipeline, movie)
```

## Rcpp Acceleration

C++ implementations are used automatically when available:

```{r rcpp}
# These functions use C++ when Rcpp is installed, R fallback otherwise

# Fast motion correction
corrected <- motion_correct_fast(movie, reference)

# Fast image operations
blurred <- gaussian_blur_fast(image, sigma = 2)
filtered <- median_filter_fast(image, size = 3)

# Fast trace extraction
traces <- extract_trace_fast(movie, mask)
```

## Benchmarking

Compare performance of different backends:

```{r benchmark}
# Generate test data
large_traces <- matrix(rnorm(1000 * 100000), nrow = 1000)

# Benchmark different approaches
benchmark_performance(
  large_traces,
  operations = c("correlation", "dff", "spike_detection"),
  backends = c("base", "data.table", "rcpp")
)

# Results show speedup factors for each backend
```

## Best Practices

### Memory Management

```{r memory}
# Check memory usage
check_memory_usage(traces)

# For large datasets, use chunked processing
corrected <- calcium_correction_chunked(
  traces,
  chunk_size = 100,  # cells per chunk
  method = "modern"
)

# Or use Arrow backend
save_traces_parquet(traces, "temp.parquet")
# Process with streaming...
```

### Choosing the Right Backend

| Dataset Size | Recommended Approach |
|--------------|---------------------|
| < 1 GB | Base R or data.table |
| 1-10 GB | data.table + parallel |
| > 10 GB | Arrow/DuckDB streaming |

### Parallel Processing Guidelines

```{r parallel-guidelines}
# Rule of thumb: workers = physical cores - 1
n_cores <- parallel::detectCores(logical = FALSE)
setup_parallel(workers = n_cores - 1)

# For memory-intensive tasks, use fewer workers
setup_parallel(workers = 4, memory_per_worker = "4GB")
```

## Session Info

```{r session, eval = TRUE}
sessionInfo()
```
