---
title: "Pharmacological Analysis for Calcium Imaging"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pharmacological Analysis for Calcium Imaging}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = FALSE
)
```

This vignette demonstrates the pharmacological analysis tools in CaImagingAnalysisFr, designed for in vitro calcium imaging experiments with drug treatments.

## Overview

When studying drug effects on cellular calcium dynamics, common questions include:

- Which cells respond to treatment?
- How does calcium activity change with drug application?
- Do cells recover after washout?
- How do calcium waves propagate through the culture?

CaImagingAnalysisFr provides specialized tools to answer these questions.

## Setup

```{r load}
library(CaImagingAnalysisFr)
```

## Experimental Design

A typical pharmacological experiment has three phases:

1. **Baseline**: Pre-drug recording

2. **Treatment**: During drug application
3. **Recovery**: After drug washout (optional)

```{r data-structure}
# Example: 10 minutes baseline, 10 minutes drug, 10 minutes washout
# At 10 Hz = 6000 frames per phase

frame_rate <- 10
baseline_frames <- 1:6000
treatment_frames <- 6001:12000
recovery_frames <- 12001:18000

# Extract phase-specific traces
baseline_traces <- traces[, baseline_frames]
treatment_traces <- traces[, treatment_frames]
recovery_traces <- traces[, recovery_frames]
```

## Responder Analysis

### Classifying Responders

Identify cells that respond to drug treatment:

```{r responders}
# Classify cells as activated, inhibited, or non-responders
resp <- classify_responders(
  baseline_traces,
  treatment_traces,
  method = "threshold",
  threshold_sd = 2
)

# View results
print(resp$summary)
# activated inhibited non_responder
#        15        8            27

# Responder fraction
print(paste("Responder fraction:", round(resp$responder_fraction * 100), "%"))

# Access per-cell results
head(resp$results)
```

### Response Subtypes

Cluster cells by their response dynamics:

```{r subtypes}
# Identify response subtypes
subtypes <- subtype_by_response(
  treatment_traces,
  n_subtypes = 4,
  method = "kmeans"
)

# View subtype characteristics
for (i in seq_along(subtypes$subtype_stats)) {
  cat(sprintf("Subtype %d: %d cells (%.1f%%), peak at frame %d\n",
              i,
              subtypes$subtype_stats[[i]]$n_cells,
              subtypes$subtype_stats[[i]]$fraction * 100,
              subtypes$subtype_stats[[i]]$peak_time))
}
```

## Baseline Calcium Analysis

### Resting Calcium Levels

Compare baseline calcium between conditions:

```{r baseline}
# Analyze baseline characteristics
baseline_analysis <- analyze_baseline(
  baseline_traces,
  method = "percentile",
  percentile = 10,
  frame_rate = frame_rate
)

print(paste("Mean baseline:", round(baseline_analysis$mean_baseline, 3)))
print(paste("Baseline stability (CV):", round(mean(baseline_analysis$stability), 3)))
```

### Calcium Load

Compute total calcium exposure (area under curve):

```{r calcium-load}
# Baseline period load
baseline_load <- compute_calcium_load(
  baseline_traces,
  frame_rate = frame_rate,
  above_baseline = TRUE
)

# Treatment period load
treatment_load <- compute_calcium_load(
  treatment_traces,
  frame_rate = frame_rate,
  above_baseline = TRUE
)

# Compare
print(paste("Baseline load:", round(baseline_load$mean_load, 2)))
print(paste("Treatment load:", round(treatment_load$mean_load, 2)))
print(paste("Fold change:", round(treatment_load$mean_load / baseline_load$mean_load, 2)))
```

## Temporal Drug Response

### Response Time Course

Characterize how the response evolves over time:

```{r temporal}
# Combine baseline and treatment for temporal analysis
full_traces <- cbind(baseline_traces, treatment_traces)
drug_onset <- ncol(baseline_traces) + 1

# Analyze temporal response
temporal <- analyze_temporal_response(
  full_traces,
  drug_onset = drug_onset,
  frame_rate = frame_rate,
  bin_size = 10  # 10-second bins
)

# Key metrics
print(paste("Time to peak:", round(temporal$time_to_peak, 1), "seconds"))
print(paste("Response type:", temporal$response_type))  # "sustained" or "transient"
```
### Response Kinetics

Fit exponential models to response dynamics:

```{r kinetics}
# Fit rise and decay kinetics
kinetics <- fit_response_kinetics(
  temporal$bin_times,
  temporal$mean_response,
  model_type = "rise_decay"
)

if (!is.null(kinetics$model)) {
  print(paste("Rise tau:", round(kinetics$tau_rise, 2), "seconds"))
  print(paste("Decay tau:", round(kinetics$tau_decay, 2), "seconds"))
}
```

## Calcium Wave Analysis

### Detecting Waves

Identify propagating calcium waves:

```{r waves}
# Requires cell positions (x, y coordinates)
# positions <- cbind(roi_centroids$x, roi_centroids$y)

waves <- detect_calcium_waves(
  treatment_traces,
  positions = positions,
  frame_rate = frame_rate,
  threshold_sd = 2,
  min_cells = 5,
  max_delay = 2  # seconds
)

print(paste("Detected", waves$n_waves, "waves"))
print(paste("Mean wave speed:", round(waves$mean_speed, 2), "um/s"))
print(paste("Wave rate:", round(waves$wave_rate, 2), "per minute"))
```

### Initiation Sites

Find cells that frequently start waves:

```{r initiation}
# Identify wave initiation sites
initiators <- identify_initiation_sites(waves, n_cells = nrow(traces), positions)

# Top initiators
head(initiators[order(-initiators$initiation_count), ], 10)
```

### Wave Participation

Measure how often each cell participates in waves:

```{r participation}
participation <- compute_wave_participation(waves, n_cells = nrow(traces))

# Cells with high participation
high_participation <- participation$highly_participating
print(paste("Cells with high wave participation:", length(high_participation)))
```

## Spatial Correlation

### Distance-Dependent Correlation

Analyze how correlation depends on cell-cell distance:

```{r spatial}
# Compute spatial correlation
spatial_baseline <- compute_spatial_correlation(baseline_traces, positions, n_bins = 10)
spatial_treatment <- compute_spatial_correlation(treatment_traces, positions, n_bins = 10)

# Space constant (correlation decay distance)
print(paste("Baseline space constant:", round(spatial_baseline$space_constant, 1), "um"))
print(paste("Treatment space constant:", round(spatial_treatment$space_constant, 1), "um"))

# Compare between conditions
comparison <- compare_spatial_correlation(spatial_baseline, spatial_treatment)
print(paste("P-value:", format(comparison$p_value, digits = 3)))
```

## Recovery Analysis

Characterize return to baseline after drug washout:

```{r recovery}
# Analyze recovery
recovery <- analyze_recovery(
  traces = cbind(baseline_traces, treatment_traces, recovery_traces),
  baseline_period = baseline_frames,
  treatment_period = treatment_frames,
  recovery_period = recovery_frames,
  frame_rate = frame_rate
)

# Recovery times
print(paste("Time to 50% recovery:", round(recovery$t50, 1), "seconds"))
print(paste("Time to 90% recovery:", round(recovery$t90, 1), "seconds"))

# Recovery classification
print(paste("Full recovery:", round(recovery$full_recovery_fraction * 100), "%"))
print(paste("Partial recovery:", round(recovery$partial_recovery_fraction * 100), "%"))
```

## Transient Shape Classification

Categorize calcium transients by their waveform:

```{r shapes}
# Classify transient shapes
shapes <- classify_transient_shapes(
  treatment_traces,
  frame_rate = frame_rate,
  n_classes = 4
)

# Shape distribution
print(shapes$shape_summary)

# Plot representative shapes
plot_transient_shapes(shapes)
```

## Excitability Metrics

Measure cellular excitability:

```{r excitability}
# Compute excitability metrics
baseline_excite <- compute_excitability(baseline_traces, frame_rate)
treatment_excite <- compute_excitability(treatment_traces, frame_rate)

# Compare event rates
print(paste("Baseline event rate:", round(baseline_excite$mean_event_rate, 2), "/min"))
print(paste("Treatment event rate:", round(treatment_excite$mean_event_rate, 2), "/min"))
```

## Comprehensive Report

Generate a full analysis report:

```{r report}
# Run comprehensive GBM analysis
report <- gbm_analysis_report(
  baseline_traces = baseline_traces,
  treatment_traces = treatment_traces,
  recovery_traces = recovery_traces,  # optional
  positions = positions,               # optional, enables spatial analysis
  frame_rate = frame_rate
)

# Print summary
print(report)

# Access specific results
report$responders$responder_fraction
report$calcium_load$treatment$mean_load
report$waves$treatment$n_waves
```

## Dose-Response Analysis

For experiments with multiple drug concentrations:

```{r dose-response}
# Example: responses at different concentrations
concentrations <- c(0, 0.1, 1, 10, 100)  # uM
responses <- c(0.05, 0.15, 0.45, 0.82, 0.95)  # normalized response

# Fit Hill equation
dr <- fit_dose_response(concentrations, responses)

print(paste("EC50:", round(dr$ec50, 2), "uM"))
print(paste("Hill coefficient:", round(dr$hill, 2)))

# Plot
plot_dose_response(dr)
```

## Statistical Comparisons

Compare conditions statistically:

```{r stats}
# Compare multiple conditions
comparison <- compare_conditions(
  list(control = control_traces, drug = treatment_traces),
  metrics = c("mean", "amplitude", "frequency"),
  test = "wilcox"
)

print(comparison)
```

## Session Info

```{r session, eval = TRUE}
sessionInfo()
```
